---
title: "StatMethodsHW1"
author: "Joshua Burkhart"
date: "January 8, 2016"
output: pdf_document
---

```{r global_options, echo=FALSE, include=FALSE, error=FALSE}
knitr::opts_chunk$set(fig.path = "Figs/",
                      message = FALSE,
                      warning = FALSE,
                      include = TRUE,
                      echo = TRUE,
                      error = TRUE,
                      fig.width = 11,
                      comment = NA)
```

```{r, echo=FALSE, include=FALSE}
library(MASS)
library(plyr) #this must be loaded before dplyr 
library(dplyr)
library(ggplot2)
library(broom)
library(knitr)
library(magrittr)
library(reshape2)
library(infotheo)
```

1. Develop an annotated R script that utilizes key exploratory data analysis techniques to identify if there are any issues in the data or potential problems for analysis.

> Load data

```{r}
data <- read.csv("~/SoftwareProjects/StatisticalMethodsInCompBio/HW1/hw1data.txt",sep="")

#copy to matrix before adding column names
data_matrix <- as.matrix(data)

# add column names to data frame
names(data) <- c(
  "numPreg",    #Number of pregnancies
  "glucTol",    #Glucose tolerance (Plasma glucose concentration @ 2 hrs)
  "diasBldPrs", #Diastolic blood pressure (mm Hg)
  "bdyFat",     #Body Fat: Triceps skin fold thickness (mm)
  "insln",      #Insulin: 2-Hour serum insulin (mu U/ml)
  "bmi",        #Body mass index (weight in kg/(height in m)^2)
  "expGenInf",  #Expected Genetic Influence of affected and unaffected relatives on subject's eventual risk
  "age",        #Age (years)
  "tstPosDbts") #Class variable (0 or 1) where 1 is interpreted as "tested positive for diabetes"
```

> Columns/Rows same size?

```{r}
for (i in 0:ncol(data)) {
    print(nrow(data[i]))
}
```

> Missing data?

```{r}
sapply(data, function(x) sum(is.na(x))) #from http://stackoverflow.com/questions/8317231/elegant-way-to-report-missing-values-in-a-data-frame
```

> Ranges?

```{r}
for (i in 1:ncol(data)) {
    print(max(data[i]) - min(data[1]))
} #Ranges differ too widely. Data should be transformed
  #onto singular range (say 0 to 1) to avoid delayed convergence of
  #gradient decent or similar convex optimization algorithms.
```

> Distributions?

```{r}
melt(data) %>% ggplot(aes(x=variable,y=value)) +
  geom_boxplot() #from http://stackoverflow.com/questions/15071334/boxplot-of-table-using-ggplot2
```

> Rank deficient/singular?

```{r}
data_inv <- ginv(data_matrix)
zapsmall(data_inv %*% data_matrix)
```

> Mutual Information?

```{r}
maxInfPos <- mutinformation(data[9],data[9])
for (i in 1:ncol(data)) {
    print(mutinformation(discretize(data[i]),data[9])/maxInfPos)
} #Column 2, Glucose tolerance (Plasma glucose concentration @ 2 hrs),
  #appears to be the best individual predictor and column 3, Diastolic
  #blood pressure (mm Hg), appears to be the worst. This may be useful
  #in narrowing features later on.
```

> Principal Component Analysis?

2. Provide any key output (figures or tables).

3. Briefly summarize any issues with the data, how you have addressed them (transforms etc) and your recommendations for further downstream analysis. Do you have any concerns about using the attribute data to predict class outcome (key here is to translated biological question->hypothesis->statistical test)?



