---
title: "StatMethodsHW2"
author: "Joshua Burkhart"
date: "January 17, 2016"
output: 
  pdf_document: 
    fig_width: 9
    fig_height: 6
    latex_engine: xelatex
---

# BMI 651: HW2

```{r global_options, echo=FALSE, include=FALSE, error=FALSE}
knitr::opts_chunk$set(fig.path = "Figs/",
                      message = FALSE,
                      warning = FALSE,
                      include = TRUE,
                      echo = TRUE,
                      error = TRUE,
                      fig.width = 11,
                      comment = NA)
```

```{r, echo=FALSE, include=FALSE}
library(MASS)
library(plyr) #this must be loaded before dplyr 
library(dplyr)
library(ggplot2)
library(broom)
library(knitr)
library(magrittr)
library(reshape2)
library(infotheo)
library(stats)
library(ggbiplot)
library(car)
```

> Based on the readings and in-class discussion,  conduct EDA and regression diagnostics to determine if there are any violations of assumptions. 

## Provide detailed script (10pts) 

> See the source .Rmd document.

## Write-up with the key figures and tables as well as description of any issues and how you handled them in order to analyze the data (20 pts) Note: you should be able to describe any output or figures without jargon!!!

> See below output.

### Load Data

```{r}
data <- read.csv("~/SoftwareProjects/StatisticalMethodsInCompBio/HW2/HIV.txt",sep="")
```

### Missing Data

```{r}
sapply(data, function(x) sum(is.na(x)))
```

> One missing ViralLoad is reported, violating an assumption of no missing or invalid data. We'll just remove it for now.

```{r}
data_rm_na <- na.omit(data)
```

### Distributions

```{r}
hist(data_rm_na$StrainScore, 
     xlab="Strain Score", 
     main="Strain Score Histogram")
```

> Unimodal wth a slight skew, looks close enough to a Gaussian...

```{r}
hist(data_rm_na$ViralLoad,
     xlab="Viral Load",
     main="Viral Load Histogram")
```

> Outcome distribution looks funny (bimodal with strong skew) and violates an assumption of normally distributed data.

### Outliers

```{r}
ssbp <- Boxplot(data_rm_na$StrainScore,
        ylab="Strain Score",
        main="Strain Score Boxplot")
ssbp #outliers reported
```

> No outliers reported for input.

```{r}
vlbp <- Boxplot(data_rm_na$ViralLoad,
        ylab="Viral Load",
        main="Viral Load Boxplot")
vlbp #outliers reported
```

> Five outliers reported in output. Row Numbers: 27, 16, 6, 21, and 43.

```{r}
model <- lm(data_rm_na$ViralLoad ~ data_rm_na$StrainScore)
summary(model)
```

> In the context of a linear model, Strain Score does not appear to be a strong predictor of Viral Load (p = 0.396).

### Leverage

```{r}
leveragePlot(model,
              term.name="data_rm_na$StrainScore",
              id.n=3,
              xlab="Strain Score (Rescaled)",
              ylab="Viral Load")
```

> The datapoints with highest leverage are from row numbers 27, 16, 24, 6, and 22.

### Influence

```{r}
plot(model, 
     which=4,
     sub.caption="(Row Number)") #plot.lm 4 is Cook's Distance
```

> We can see that data from rows 16 and 27 appear highly influential.

```{r}
cdbp <- Boxplot(cooks.distance(model))
cdbp
```

> This boxplot presents the distribution of influential points and labels data from rows 2, 5, 6, 9, 16, 19, 24, 27, 43 as outliers.

### Model

```{r}
plot(data_rm_na$StrainScore,data_rm_na$ViralLoad,
     xlab="Strain Score",
     ylab="Viral Load", 
     main="Viral Load ~ Strain Score (p = 0.396)")
abline(model, col="red")
```

> The model does appear to fit the data but we must remember that it does not do so significantly.

### Rescue?

```{r}
data_rm_na[27,] 
range(data_rm_na[-27,][2])
```

> Data Row 27 reports a Viral Load almost triple the range for all other points. If we assume this is a mistake, we can remove it and rerun our model.

```{r}
data_rm_na_no27 <- data_rm_na[-27,]
model_no27 <- lm(data_rm_na_no27$ViralLoad ~ data_rm_na_no27$StrainScore)
summary(model_no27)
```

> Strain Score becomes a significant predictor of Viral Load after removing row 27.

```{r}
plot(data_rm_na_no27$StrainScore,data_rm_na_no27$ViralLoad,
     xlab="Strain Score",
     ylab="Viral Load", 
     main="Viral Load ~ Strain Score (p = 0.0378)")
abline(model_no27, col="red")
```

> We must be cautious in accepting this result as row 27 may not have been a mistake after all. Viral Loads over 1 million have been reported[1].

## References

[1] http://www.catie.ca/en/fact-sheets/transmission/hiv-viral-load-hiv-treatment-and-sexual-hiv-transmission
