---
title: "StatMethodsHW3A"
author: "Joshua Burkhart"
date: "January 26, 2016"
output: 
  pdf_document: 
    fig_width: 9
    fig_height: 6
    latex_engine: xelatex
---

# BMI 651: HW3 A

```{r global_options, echo=FALSE, include=FALSE, error=FALSE}
knitr::opts_chunk$set(fig.path = "Figs/",
                      message = FALSE,
                      warning = FALSE,
                      include = TRUE,
                      echo = TRUE,
                      error = TRUE,
                      fig.width = 11,
                      comment = NA)
```

```{r, echo=FALSE, include=FALSE}
library(MASS)
library(plyr) #this must be loaded before dplyr 
library(dplyr)
library(ggplot2)
library(broom)
library(knitr)
library(magrittr)
library(reshape2)
library(infotheo)
library(stats)
library(ggbiplot)
library(car)
library("Hiiragi2013")
```

## (1) draft analysis plan for developing classifier for predating developmental stage (E3.25) based on gene expression

> 1. Check/correct data typing. Ex: factors encoded as ints  
  2. Check/remove rows with missing data  
  3. Check/remove rows with nonsense data
  4. Check/Z-score ranges / distributions
  4. Randomly split data into training / test sets
  5. see http://www.r-bloggers.com/fitting-a-neural-network-in-r-neuralnet-package/  
  6. PCA test set? http://stats.stackexchange.com/questions/121522/how-to-transform-test-set-to-the-pca-space-of-the-training-set-if-the-features  

## (2) Write R script for performing EDA on the data set and then perform EDA

## (3) Create version 2 of analysis plan that includes EDA and any revisions (if needed) based on EDA 

## (4) Write R script for analyzing the data 

## For submission A, provide the 2 analysis plans and 2 R scripts. 
