---
title: "StatMethodsHW3B"
author: "Joshua Burkhart"
date: "February 3, 2016"
output: 
  pdf_document: 
    fig_width: 9
    fig_height: 6
    latex_engine: xelatex
---
# BMI 651: HW3 B

```{r global_options, echo=FALSE, include=FALSE, error=FALSE}
knitr::opts_chunk$set(fig.path = "Figs/",
                      message = FALSE,
                      warning = FALSE,
                      include = TRUE,
                      echo = TRUE,
                      error = TRUE,
                      fig.width = 11,
                      comment = NA)
```

```{r, echo=FALSE, include=FALSE}
library(MASS)
library(plyr) #this must be loaded before dplyr 
library(dplyr)
library(ggplot2)
library(broom)
library(knitr)
library(magrittr)
library(reshape2)
library(infotheo)
library(stats)
library(ggbiplot)
library(car)
library("Hiiragi2013")
set.seed(2013) #does this need to be set to 2013? why?
```
  
### Load Data
  
```{r}
data("x")
data("xq")

#feature candidates
hw3A.x.genotypes <- x@phenoData@data$genotype
hw3A.x.probe_intensities <- data.frame(assayDataElement(x@assayData,'exprs'))
hw3A.xq.cell_type <- xq@phenoData@data$Cell.type
hw3A.xq.gene_expressions <- data.frame(assayDataElement(xq@assayData,'exprs'))

#classes
hw3A.x.classes <- x@phenoData@data$Embryonic.day
hw3A.xq.classes <- xq@phenoData@data$Embryonic.day
```

### Scan for missing Data

```{r}
sum(is.na(hw3A.x.genotypes))
sum(is.na(hw3A.x.probe_intensities))
sum(is.na(hw3A.xq.cell_type))
sum(is.na(hw3A.xq.gene_expressions))

sum(is.na(hw3A.x.classes))
sum(is.na(hw3A.xq.classes))
```

```{r}
t(hw3A.xq.gene_expressions) %>% summary()
hw3A.xq.gene_expressions <- na.omit(hw3A.xq.gene_expressions)
```

### Transform positive (E3.25) and negative (E3.5 and E4.5) values 1 and 0, respectively

```{r}
hw3A.x.classes <- as.character(hw3A.x.classes)
hw3A.x.classes[hw3A.x.classes == "E3.25"] = 1
hw3A.x.classes[hw3A.x.classes == "E3.5"] = 0
hw3A.x.classes[hw3A.x.classes == "E4.5"] = 0
hw3A.x.classes <- as.factor(hw3A.x.classes)

hw3A.xq.classes <- as.character(hw3A.xq.classes)
hw3A.xq.classes[hw3A.xq.classes == "E3.25"] = 1
hw3A.xq.classes[hw3A.xq.classes == "E3.5"] = 0
hw3A.xq.classes[hw3A.xq.classes == "E4.5"] = 0
hw3A.xq.classes <- as.factor(hw3A.xq.classes)
```

### Combine Features for x and xq datasets

```{r}
hw3A.x.features <- rbind(hw3A.x.genotypes,hw3A.x.probe_intensities)
hw3A.xq.features <- rbind(as.factor(hw3A.xq.cell_type),hw3A.xq.gene_expressions)
```

### Split into Training & Test sets

```{r}
# x
index <- sample(1:ncol(hw3A.x.features),round(0.8*ncol(hw3A.x.features)))
hw3A.x.features_train <- hw3A.x.features[,index]
hw3A.x.classes_train <- hw3A.x.classes[index]
hw3A.x.features_test <- hw3A.x.features[,-index]
hw3A.x.classes_test <- hw3A.x.classes[-index]

# xq
index <- sample(1:ncol(hw3A.xq.features),round(0.8*ncol(hw3A.xq.features)))
hw3A.xq.features_train <- hw3A.xq.features[,index]
hw3A.xq.classes_train <- hw3A.xq.classes[index]
hw3A.xq.features_test <- hw3A.xq.features[,-index]
hw3A.xq.classes_test <- hw3A.xq.classes[-index]
```

### Z-score training set

```{r}
# x
x.train_sd = matrix()
x.train_mean = matrix()
hw3A.x.features_train_normalized <- hw3A.x.features_train
for(i in 2:nrow(hw3A.x.features_train)) # each row is a feature, first is a factor (genotype)
  x.train_sd[i] = sd(hw3A.x.features_train[i,])
  x.train_mean[i] = apply(hw3A.x.features_train[i,],1,mean)
  for(j in 1:ncol(hw3A.x.features_train)) # each column is a sample
    hw3A.x.features_train_normalized[i,j] =
    (hw3A.x.features_train[i,j] - x.train_mean[i]) / x.train_sd[i]

# xq  
xq.train_sd = matrix()
xq.train_mean = matrix()
hw3A.xq.features_train_normalized <- hw3A.xq.features_train
for(i in 2:nrow(hw3A.xq.features_train)) # each row is a feature, first is a factor (cell_type)
  xq.train_sd[i] = sd(hw3A.xq.features_train[i,])
  xq.train_mean[i] = apply(hw3A.xq.features_train[i,],1,mean)
  for(j in 1:ncol(hw3A.xq.features_train)) # each column is a sample
    hw3A.xq.features_train_normalized[i,j] =
    (hw3A.xq.features_train[i,j] - xq.train_mean[i]) / xq.train_sd[i]
```

> With our datasets now split and Z scored, we are now ready to move on to feature selection (in the next assignment).

## Feature Selection (Part B)

### Wilcoxan Rank Sum

> First we'll attempt to reduce our prospective features using the wilcoxan rank sum test and a p-value limit of 0.01. We'll store the indices of the features we discover so we can filter any new test data in the same way.

```{r}
P_LIM = 0.01
x.train_w_idx = vector()
xq.train_w_idx = vector()

if(FALSE){ # takes too long.. skip this for now
#x
for (i in 1:nrow(hw3A.x.features_train_normalized)){
  x.w <- wilcox.test(unlist(hw3A.x.features_train_normalized[i,]) ~ hw3A.x.classes_train,
                     data = hw3A.x.features_train_normalized)
  if(x.w$p.value < P_LIM)
    x.train_w_idx <- append(x.train_w_idx,i) #store indices of significant features
  print(x.train_w_idx %>% length())
}
  
}

#xq
for (i in 1:nrow(hw3A.xq.features_train_normalized)){
  xq.w <- wilcox.test(unlist(hw3A.xq.features_train_normalized[i,]) ~ hw3A.xq.classes_train,
                     data = hw3A.xq.features_train_normalized)
  if(xq.w$p.value < P_LIM)
    xq.train_w_idx <- append(xq.train_w_idx,i) #store indices of significant features
  print(xq.train_w_idx %>% length())
}
```

## References

> [1] http://www.r-bloggers.com/fitting-a-neural-network-in-r-neuralnet-package/  
  [2] http://stats.stackexchange.com/questions/121522/how-to-transform-test-set-to-the-pca-space-of-the-training-set-if-the-features  
  [3] https://support.bioconductor.org/p/40730/  
  [4] https://bioconductor.org/packages/release/data/experiment/manuals/Hiiragi2013/man/Hiiragi2013.pdf  
  [5] http://127.0.0.1:17064/library/Hiiragi2013/doc/Hiiragi2013.pdf  
  [6] http://www.rdatamining.com/examples/decision-tree  
